Developing the algebraic hierarchy with type classes in Coq

### 2010-02-25 roughdiamond.tex
first commit

\documentclass[a4paper,10pt, runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{ucs}
% \usepackage[utf8x,utf8]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs%i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Category theory provides a further
useful abstraction. We provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions. Similarly, we provide an interface for polynomials over a ring $R$ using $R$-algebras.
We show that the standard implementations of these are in fact instances of these type classes.
We report on the efficiency penality of these abstractions.

Finally, we use type classes to conveniently implement a quote function using the prolog-like
abilities of type class unification. We integrate this with our development of universal algebra.
\end{abstract}

\section{Introduction}
Developing formalized libraries for mathematics requires a detailed analysis of the
architecture of mathematics~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large
software project.
The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
inheritance: A ring is a type with two operations. How do we inherit from the two theories of
monoids? 2. Sharing of notations. 3. Combining syntactic and semantic methods, usually this is
addressed by reflection.

Several solutions have been proposed for the Coq theorem prover~\cite{Coq,BC04}: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction to
OCaml~\cite{letouzey02}. Modules may also be seen as parametrised sections. They allow management
of namespaces, notations, hints, etc. An |Import| command imports all the constants from a
module. Thus allowing the user to type |B| instead of |A.B|. Similar functionality would be useful
for type classes.}
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

In this fashion we implement\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}}
 algebraic theories as type classes, not instances, thus addressing the
challenges above; see Section~\ref{classes}. Type classes have recently entered
Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq,
type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.

We abstract from concrete algebra and formalize many-sorted universal algebra, including a
construction of
free $\Sigma$-algebras; see Section~\ref{univ}. We are also concerned with efficiency.
In fact, our aim is to use the present development as a new basis for Bishop's program in type
theory~\cite{typesreal-article}.\marginpar{More}

So, we would like to replace data structures with more efficient ones. Concretely,
Coq provides unary, binary and machine numbers. We use type classes polymorphism to
develop a unified library for them; see Section~\ref{modul}.
The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of
fractions; see Section~\ref{interfaces}. In future work we aim to base our development of the reals
on an abstract dense set, allowing us to use the efficient dyadic
rationals~\cite{boldo2009combining} as a base for exact real number computation in
Coq~\cite{Riemann,Oconnor:real}.

The paper is organised as follows: Section~\ref{classes} starts with a development of the concrete
algebraic hierarchy addressing the issues raised above. Section~\ref{univ} abstracts to universal
algebra, we develop the basic theory to the point of the first homomorphism theorem.
Section~\ref{interfaces} provides interfaces to usual datastructures by providing their universal
properties. To allow symbolic manipulation of semantic objects one needs a quote function. Usually,
this function is implemented in Coq's tactics language. A more convenient way is to use type class
unification; see section~\ref{quote}. In section~\ref{canonical}, we end with a short comparision
between type classes and canonical structures.

In this paper we focus on the Coq proof assistant. We conjecture that the methods can be transfered
to any type theory based proof assistant supporting type classes such as
Matita~\cite{asperti2007user}.

\section{Preliminaries}
\subsection{Coq type theory}
The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}.\marginpar{More}
This type theory lacks quotient types, as it would make type checking undecidable. Instead it
supports setoids (aka Bishop sets), a pair of a type and an equivalence
relation~\cite{Bishop67,Hofmann,Capretta}. Recently, improved support has been added to the
system~\cite{Setoid-rewrite}. Functions respecting these equivalence relations are called |Proper|.
As we will see in Section~\ref{univ2} below this functionality can also be used to work with
quotient algebras.

\subsection{Type classes}
Implementation in Ltac, Notation etc
Coercions

\section{To bunde or not to bundle? (Or: What to bundle?)}

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
%format → = "\rightarrow"
%format Π = "\Pi"
\begin{code}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}
and similar for transitive and symmetric relations.

With this approach, we run into problems when we try to define |EquivalenceRelation| in terms of
|ReflexiveRelation| and its two siblings, because when posit three objects of type
|ReflexiveRelation|, |TransitiveRelation|, and |SymmetricRelation|, respectively, we have to
add equalities expressing that they talk about the same carrier and relation, and these equalities
are a nightmare to work with (because the three carriers and relations won't be convertible).
Technically, this leads to the research in manifest fields, the \emph{with} construction, to which
we return in Section~\ref{manifest}.

As a first step we would define |ReflexiveRelation| as:
\begin{code}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: rr_carrier → rr_carrier → Prop
    ; rr_proof: Π x, rr_rel x x }.
\end{code}

This would solve the problem for the carrier, but |EquivalenceRelation| would still need equalities
for the relation:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_refl: ReflexiveRelation er_carrier
    ; er_sym: SymmetricRelation er_carrier
    ; er_trans: TransitiveRelation er_trans
    ; same_rel: rr_rel er_refl = sr_rel er_sym
    ; same_rel': rr_rel er_refl = tr_rel er_trans }.
\end{code}

A natural solution is to pull the relation out of |ReflexiveRelation| (and its siblings) as well:
\begin{code}
  Record ReflexiveRelation (carrier: Type)
                           (rel: carrier → carrier → Prop): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{code}
or equivalently:
\begin{code}
  Definition ReflexiveRelation (carrier: Type)
                               (rel: carrier → carrier → Prop): Prop :=
    Π x, rr_rel x x.
\end{code}
This is how reflexivity is defined in the standard library, and this is entirely sensible. The
change from |Type| to |Prop| reflects the fact that |ReflexiveRelation| is now strictly a property,
proofs of which can be nicely kept opaque;%
\footnote{Coq distinguishes between transparent and opague definitions. The former can be unfolded,
whereas this is impossible for the latter.}
we will return to this issue in Section~\ref{explicit}.
We can now define |EquivalenceRelation| without awkward equalities:
\begin{code}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: er_carrier → er_carrier → Prop
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

But looking at |EquivalenceRelation|, we see that it now essentially has the same form as the
original naive |ReflexiveRelation| record we started with. Clearly, equivalence relations
and reflexive relations are analogous concepts, and by the same argument we
conclude that |EquivalenceRelation| should really be defined as:

\begin{code}
Record EquivalenceRelation (er_carrier: Type)
                             (er_rel: er_carrier → er_carrier → Prop): Prop :=
  { er_refl: ReflexiveRelation er_carrier er_rel
  ; er_sym: SymmetricRelation er_carrier er_rel
  ; er_trans: TransitiveRelation er_trans er_rel }.
\end{code}

After all, when we define a type with
two different equivalence relations on it,
such as the quotient of a setoid,
we have the same considerations as for |ReflexiveRelation| above.

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of |Monoid| that bundles the
relation. Then if we proceed to define |SemiRing| as a double |Monoid|, we face a dilemma:
\begin{code}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, ?
\end{code}
Which equality do we use in the expression ?, |mult_monoid|'s equality or |plus_monoid|'s?
Again, for this definition to be usable at all, we would need to add an equality expressing that the
two monoids contain the same equivalence relation. And again, the proper solution is to change
Monoid and unbundle the relation. But here, we have a choice as to how much we unbundle exactly.
(For the moment we ignore the question of whether the unit and binary operator should be unbundled
as well, focusing only on the equivalence relation and its equivalence proof.)

We \emph{could} become over-eager and decide to unbundle the equivalence proof term as well:
\begin{code}
Record Monoid (A: Type) (R: A → A → Prop)
  (e: Equivalence R) : Type := ...
\end{code}
However, this way if we happen to have two proofs |p| and |q| that a certain relation is an
equivalence, we end up with two non-convertible ways to express that the relation (together with
some unit and binary operation) form a Monoid: 'Monoid A R p ...' and 'Monoid A R q ...'. We must
either try to deal with this (by proving that Monoid proofs with different Equivalence proofs are
interchangable, and hope for automation, such as type class instance resolution, to use them
interchangably), or take great care to ensure that |Equivalence| proof terms are always shared,
essentially elevating them to `values', which goes against the principle of proof
irrelevance.\marginpar{What if we add proof irrelevance?}

Neither of these options are very appealing. We really should just make |Monoid| take as parameters
only the operations/relations. If we treat the monoid's unit and binary operation the same way, this
means that |Monoid| ends up in |Prop|, and that its instances can be left opaque, which is really
how things ought to be.

Another benefit of this approach is that it works very smoothly when these records are in fact type
classes. When relations and operations are bundled and do \emph{not} appear as parameters of the
structure record/class, expressing as a type class constraint that some collection of relations and
operations posess the structure seems to require manifest fields; see Section~\ref{manifest}. With
our unbundled approach, on the other hand, such type class constraints are easily expressed without
any extensions as simple type inhabitation queries
% (e.g. ``hey Coq, find me a proof of [Monoid some_rel some_op some_unit]'')
of types living in Prop, so that we don't even care what instances are found (thus embracing proof
irrelevance).

We remark that leaving the equivalence proofs bundled means that when one has a proof that something
is a semiring, one has two proofs (via the two monoids) that the equivalence relation is an
equivalence. Fortunately, having redundant proofs like this is entirely harmless when one's terms
never refer (directly or indirectly) to proofs, which is the case when the relations and operations
they are composed of are left unbundled. Here, too, we enjoy the warm embrace of proof irrelevance.

From the discussion so far, we derive some principles:
\begin{enumerate}
 \item structural concepts (like Reflexive, Setoid, Monoid) should have their structural components
(i.e. carriers, relations, and operations) as parameters, and nothing else;
 \item (as a consequence) these concepts should be defined as predicates (which includes records and
classes living in Prop), and their proofs may be kept opaque;
 \item (as a second consequence) algebraic expressions need/should never refer (directly or
indirectly) to proofs.
\end{enumerate}
% \setcounter{enumi}{3}

This list will grow in subsequent sections.

The principles listed so far are not particularly controversial, in fact this
approach ensures maximum flexibility~\cite{Hints}. However,
there is a common perception that unbundling carriers, relations, and operations this way invariably
leads either to unmanagably big algebraic expressions (involving endless projections and/or implicit
arguments), or to unmanageably long and tedious enumeration of carriers/relations/operations, or to
problems establishing canonical names for operations (since we no longer project them out of
structure instances all the time).

In the next sections, we show how systematic use of type classes (and their support infrastructure
in Coq), combined with the use of a special form of type class we call an `operational type class'
(as opposed to a `structural' type class we might use in place of the records shown thus far), lets
us avoid these problems.

\section{Managing unbundled structure with type classes}

So far, we have only looked at the definitions of the structures themselves, and have not yet
considered the perspective of someone trying to prove some theorems about these abstract structures.
We would like to approximate the common mathematical vernacular, where one
simply says:

for $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$, $\ldots$

Suppose we have defined |SemiGroup| in line with the aforementioned principles:
\begin{code}
Record SemiGroup (A: Type) (eq: A → A → Prop)
 (op: A → A → A): Prop := ...
\end{code}
Then our starting point is the obvious but awful:
\begin{code}
Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                   eq (op (op x y) z) (op (op z y) x).
\end{code}
As a first cleaning measure, we might hope to introduce some notations to let us use infix notation
for the semigroup operator (say, |&|) and for equality. Unfortunately, at this point the names |eq|
and |op| are just local names and there is no way to predeclare that they are to be associated with
any notation. What we really need are canonical names for these notions, so that we may fashion
notations for them.

This is a typical job for type classes, and they are used in this way in the |Clean| standard
library~\cite{Clean}. Similarly, Haskell uses type classes to overload notations.
We simply follow suit:
\begin{code}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A → A → A.

  Infix "=" := equiv: type_scope.
  Infix "&" := sg_op (at level 50, left associativity).
\end{code}
We choose to use |&| here and reserve the notation |*| for ring multiplication.

We use = for setoid equality, since Leibniz equality seems less important for the formalization of
mathematics. More precisely, setoid equality is the basic concept. In special case, one can prove
that setoid equality coincides with Leibniz equality.

We would have like to use a haskell style definition
\begin{code}
Class Equiv A := (=): relation A.
\end{code}
Unfortunately, currently this is not supported in Coq.

These single-field type classes are given special treatment in that they are not translated into
records the way classes with any other number of fields are, but are instead turned into ordinary
definitions. The effect of this is that |Equiv A| immediately reduces to |relation A|, and that the
equiv `projection' is the identity function whose argument is inferred using instance resolution.
Thus, for a very modest cost in indirection, we gain the ability to use canonical names, and with
it, notations:
\begin{code}
  Record SemiGroup (A: Type) (eq: Equiv A)
                             (op: SemiGroupOp A): Prop := ...
  Lemma assoc1 A eq op (P: SemiGroup A eq op) x y z:
                    x & y & z = z & y & x.
\end{code}
We call these type classes \emph{operational type classes}. We note that operational type classes
allow us to avoid Coq's notation |scope| mechanism.

It is important to point out that although the expression now \emph{looks} like it is bound to some
specific semigroup structure (which with traditional approaches would imply projections from bundles
and/or reference to proofs), it is really only referring near-directly to the actual operations
involved, with the semigroup proof existing only as opaque `knowledge about the operations' which we
may use in the proof. This lack of projections keeps our terms small and independent, and keeps
rewriting simple and sane.

On the other hand, it \emph{does} mean that all the carriers/relations/operations end up in the
context. The context for |assoc1| looks like:
\begin{code}
  A: Type
  eq: Equiv A
  op: SemiGroupOp A
  P: SemiGroup A eq op
\end{code}

While we are not particularly worried about overly large contexts (we feel that this will most
likely not be problematic for any but the most complex formalizations), having to \emph{declare}
these entities (like |eq| and |op| for |assoc1|) \emph{is} a bit of a chore. Fortunately, Coq
provides a feature called implicit generalization, which is exactly what we need.
\marginpar{describe how it works} With it, we can write |assoc1| as follows:
\begin{code}
  Lemma assoc1 `(SemiGroup A) x y z: x & y & z = z & y & x.
\end{code}
Thus, we have reached the mathematical ideal we aimed for.

Let us now stop and think to what extent this approach suffers from all the problems commonly
associated with it. In particular, let us imagine what happens to our terms and contexts when we
want to talk about nested structures such as polynomials over the ring of matrices over a
given ring; see Section~\ref{interfaces}.

\section{Subset types}
BEGIN SPECULATIVE

A record type is \marginpar{mostly?} syntactic sugar for a $\Sigma$-type. In the presence of a
squash operator, |[]:Type->Prop|, we may define a subset type as |\{A\mid B\}=Sig A. [B]|. This is
pursued by Barras~\cite{Barras:subset,Werner}. The program machinery fakes subset types in the
present Coq implementation. Proof irrelevanve in Coq would allow a faithful
implementation.

Squash is a monad and allows many constructions: subsets, unions, images, heterogeneous equality,
\ldots.

Question: Would the meta-theory simplify if we would just add a conversion rule $x\triangleright
\epsilon$ for $x:Prop$? We could add a term constructor $[a]$ which will have type $[A]$. There
seems to be no reason to connect this to |Prop| or impredicativity.

END SPECULATIVE

We add another principle:
\begin{itemize}\setcounter{enumi}{3}
 \item Working hypothesis: `every record can be reordered as a subset'. Consequently, we
rewrite this subset |\{A\mid B\}| as |Record r (a:A):={b:B}|.
\end{itemize}

\section{Algebraic hierarchy using type classes}\label{classes}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.
The notation |:>| makes |semiring_mult_monoid| a coercion: when asked for a monoid it suffices to
provide a ring.

@@

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively
harmless\footnote{Coq's data structure for contexts is not very efficient however, we believe this
may be a bottleneck for huge developments. It seems that the data structure chosen
in~\cite{asperti2009compact} will behave better.}%
: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

\subsection{Manifest fields}\label{manifest}

@@

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalized some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. It has recently been extended with more
categorical results~\cite{dominguez2008formalizing}. We have conveniently redeveloped most of this
using type classes. We defined equational theories and their varieties:
the categories of their models. A quote function helps to translate a statement for an instance of
a type class to a statement in the equational logic.

\subsection{Category theory}\label{cats}
Using type classes, we provide a basic library for category theory: categories, functors,
adjunctions, monads,\ldots We follow the usual type theoretical definition of a\marginpar{Use
notation for comp} category~\cite{saibi1995constructive}
\begin{code}
 Class Category O `{!Arrows O} `{forall x y: O, Equiv (x --> y)}
   `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> forall x y: O, Setoid (x --> y)
  ; comp_proper:> forall x y z: O,
      Proper (equiv ==> equiv ==> equiv)%signature (@comp _ _ _ x y z)
  ; comp_assoc: forall (w x y z: O) (a: w --> x) (b: x --> y) (c: y --> z),
      comp c (comp b a) == comp (comp c b) a
  ; id_l: forall (x y: O) (a: y --> x), comp cat_id a == a
  ; id_r: forall (x y: O) (a: x --> y), comp a cat_id == a }.
\end{code}
This is based on the 2-categorical idea that we have equality on arrows, but not on objects.
Similarly, we have equality on natural transformations, but not on functors.

%format ◎ = "\circ"
For instance natural transformations are formalized as: |◎|
\begin{code}
  Context `{Category C} `{Category D} (F G: C -> D)
     `{!Functor F Fa} `{!Functor G Ga}.

  Class NaturalTransformation (η: forall c: C, F c --> G c): Prop :=
    natural: forall (x y: C) (f: x --> y), η y ◎ fmap F f == fmap G f ◎ η x.
\end{code}

The notation |◎| denotes function composition.
Here |fmap F| explicitly refers to the arrow part of the functor |F|.\\
In mathematics one would apply $F$ to both objects and arrows ($F x$, $F f$). However, this does
not seem to fit with type theory, hence we follow the customs in functional programming.

The proofs are short and follow the textbook easily.\marginpar{really?}

We provide the category of setoids, the category of categories, the dual of a category, products,
and hence co-products, the category of $\Sigma$-algebras.

We provide the construction of the term-algebra for a signature $\Sigma$ and show that it is
initial in the category of $\Sigma$-algebras. We proceed to prove that this provides a left-adjoint
to the forgetful functor. The resulting expression monad will be used below.

\subsection{Universal algebra continued}\label{univ2}

We have developed basic universal algebra~\cite{meinke1993universal} upto the first homomorphism
theorem.~\marginpar{Description of the work.}
%perhaps the soundness theorem for equational logic.

In order to prove the first homophism theorem, we define a congruence:
%theory/ua_congruence.v
\begin{code}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{code}
We note the use of |Proper|.

We define the forgetful functor from the category of $\Sigma$-algebras to the category of sets.

We define the term algebra consisting of closed terms, terms with the empty type as set of
variables. % extensionally equal closed term???
We show that the term algebra is initial. One may expect that this could be useful to
\emph{define}, say, the natural numbers as the initial semiring. However, it seems difficult to
prove decidability of equality, as it seems to require a normalization procedure.

We define the product of two algebras and show that it is the categorical product.

Given a sub\emph{set} of an algebra, we define the sub\emph{algebra} generated by it.
An interesting fact to mention is the use of the following heterogeneous equality between
elements of the algebra and of the subalgebra, i.e.\ terms of different types may be equal.
\begin{code}
  Fixpoint heq {o}: op_type carrier o -> op_type v o -> Prop :=
    match o with
    | constant x => fun a b => `a == b
    | function x y => fun a b => forall u, heq (a u) (b (`u))
    end.
\end{code}

We connect the abstract theory to the concrete theory. Concretely, let $\Sigma$ be the theory
of monoids. Given a monoid, we can construct the corresponding $\Sigma$-algebra and an
object in the category of $\Sigma$-algebras. Conversely, given an object in the category, we
construct a monoid.
\begin{code}
 Variable o: Variety theory.
 Global Instance from_object: Monoid (o tt).
\end{code}
Similarly, for arrows.

%There is a slight problem when defining the forgetful functor from the category of
%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
%categories\footnote{Perhaps we should have used higher categories for this.}.
%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
%definitions.
\section{Interfaces using category theory}\label{interfaces}\label{modul}
The current work was stimulated by our desire~\cite{Riemann} to move between several
implementations in a flexible way. In haskell type class polymorphism is used for this purpose.
Hence, we do the same in type theory.

We have characterized the naturals as the initial object in the category of semirings and derived
many of their properties from this interface. Unary, binary and machine numbers are
instances of this interface, so we can directly apply these results to all these instances.
Similarly, for the integers are the initial ring. The rationals are the field of fractions of the
integers. 
Given a ring $R$, the $R$-algebra $R[X]$ of polynomials, is the free $R$-algebra of a set $X$.
We provide two implementations of polynomials: the
standard representation using lists of coefficients and the Bernstein representation. A Bernstein
basis polynomial is one of the form:
\[b_{\nu,n}(x) = {n \choose \nu} x^{\nu} \left( 1 - x \right)^{n - \nu}, \quad \nu = 0, \ldots, n.\]
A Bernstein polynomial is a linear combination of these basic polynomials. Bernstein polynomials
have been used for efficient computations inside the Coq system~\cite{ZumkellerPhD}.

We encourage more efficient implementations by assigning the default implementation a
low priority. For example, the distance function on the natural numbers, which is derived from its
semiring structure, is assigned priority 10.
\begin{code}
  Global Program Instance: NatDistance N | 10 := ...
\end{code}

The context grows as follows:\marginpar{expand}

\section{Type class quoting}\label{quote}
Unification hints~\cite{Hints} allow one to semi-automatically construct a quote
function\footnote{Gonthier provides similar functionality using canonical structures}.
Unification hints are absent in Coq. Fortunately, type classes provide similar functionality as
we will now show.

We define a term language.
\begin{code}
Inductive Expr (V: Type) := Mult (a b: Expr V) | Zero | Var (v: V).
\end{code}
This defines a monad on Type\marginpar{Do this!}.
The expression type is parameterized over the set of variable indices. Hence, we diverge
 from~\cite{Hints}, which uses |nat| for variables thereby introducing bounds problems and
 dummy variables.

\noindent An expression is only meaningful in the context of a variable assignment:\footnote{nat is
an Expr-algebra. eval is map vars composed with the algebra map.}
\begin{code}
Definition Value := nat.
Definition Vars V := V -> Value.

Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
  match e with
  | Zero => 0
  | Mult a b => eval vs a * eval vs b
  | Var v => vs v
  end.
\end{code}
%
Monads are trees with grafting. As explained by sigfpe~\cite{MonadsGrafting}, the tree monad
$TX:=1+T^2X$, may be seen as the prototypical monad. Indeed, the |Expr| monad, i.e.\ the free
$\Sigma$-algebra monad, behaves much like the tree monad. We now provide a direct definition of the
corresponding bind operation.

\begin{code}
Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
  match e with
  | Zero => Zero
  | Mult a b => Mult (bind f a) (bind f b)
  | Var v => vs v
  end.
\end{code}

We have shown that |Expr|, i.e.\ |bind| together with |Var|, is a monad on the category of Types
with extensional functions between them~\marginpar{Really?}.

>>>>

\documentclass[a4paper,10pt]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage[utf8x]{inputenc}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
%include coq2tex.tex
\arrayhs %i.e. no page breaks in code blocks
\divide\abovedisplayskip 4
\divide\belowdisplayskip 4
\begin{document}
\title{Developing the algebraic hierarchy with type classes in Coq}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\maketitle
\begin{abstract}
We propose generic design patterns to organize algebraic structures using type classes inside the
Coq system. We support multiple inheritance, sharing of notations and theories, and automated
structure inference. A library for many-sorted universal algebra abstracts from the specific
algebraic structures and aids the organization of the library. Finally, we provide abstract
interfaces for $\N$, $\Z$, $\Q$ as the initial semiring, the initial ring and a field of
fractions.
\end{abstract}

\section{Introduction}
Developing formalized libraries for mathematics requires a detailed analysis of the architecture~\cite{C-corn,DBLP:conf/types/HaftmannW08}, similar to that of a large software project. 
The development of an algebraic hierarchy is an important aspect. Challenges include: 1. multiple
inheritance: A ring is a type with two operations. How do we inherit from the two theories of
monoids? 2. Sharing of notations.

Several solutions have been proposed for the Coq theorem prover%~\cite{Coq,BC04}
: dependent
records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a.\ telescopes), packed classes~\cite{Packed} and
occasionally modules. Although the latter option is quickly discarded%
\footnote{Modules are not first class, but currently they seem to be superior for extraction
purposes%~\cite{letouzey02}%
.}, considering
algebraic theories as `interfaces', as opposed to `modules', has been an important
guiding principle~\cite{Packed}.

In this fashion we implement\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
 algebraic theories as type classes, not instances, thus addressing the
challenges above; see Section~\ref{classes}. Type classes have recently entered Coq~\cite{DBLP:conf/tphol/SozeauO08}. In Coq
type classes are \emph{first class} citizens based on dependent records. The presence of dependent types makes them even more
powerful than the already succesful second class implementations in Haskell and Isabelle. Moreover,
Coq's type classes are well integrated with setoid-rewriting~\cite{Setoid-rewrite}.

% In
% Coq, type classes have advantages over modules\footnote{
% Modules are sections on steroids: namespace management, notations, hints, etc.
% The [Import] command imports all the constants from a module. Thus allowing the user to type B
% instead of A.B
% }:
% they are first class and extract~\cite{letouzey02} to Haskell\footnote{
% Would module like extraction be possible?}.

We abstract from concrete algebra and formalize many-sorted universal algebra, including a construction of
free $\Sigma$-algebras; Section~\ref{univ}. Finally, we are concerned with efficiency. So, we would like to replace data structures with more efficient ones. For instance, 
Coq provides unary, binary and machine numbers. Hence, we use ad-hoc polymorphism with type classes to
develop a unified library for them; see Section~\ref{modul}. The interfaces for $\N$, $\Z,\Q$ are the initial semiring, the initial ring and a field of fractions; see Section~\ref{interfaces}.

%Our contributions are:\footnote{The sources are available at:~\url{http://www.eelis.net/research/math-classes/}}%
%\begin{itemize}\itemsep=0pt
% \item The implementation of an algebraic hierarchy based on type classes, addressing the
%issues raised above; see Section~\ref{classes}.
% \item A formalization of many-sorted universal algebra in Coq, including a construction of
%free $\Sigma$-algebras; Section~\ref{univ}.
%\end{itemize}

\section{Algebraic hierarchy using type classes}\label{classes}
\vspace*{-2ex}
\paragraph{Canonical names.}\label{names}
Sharing of notations is an important aspect of the design of a mathematical library. Haskell uses
type classes to overload notations. In Coq we assign a class to every operation, thus avoiding Coq's notation |scope| mechanism.
 Each such operation class has only \emph{one} constructor, hence no record is constructed.
\begin{code}
 Class Equiv A := equiv: relation A.
 Infix "==" := equiv (at level 70, no associativity).
\end{code}
% Instance: Params (@equiv) 2.
\vspace*{-5ex}
\paragraph{Hierarchy.}\label{hierarchy}
In our formalization of algebraic structures we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{code}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{code}
where e.g.\ |RingPlus| is the operation class for the semigroup operator.

@@

We mention the trade-off between bigger contexts versus bigger terms. Our contexts are bigger than
those of telescopes or packed classes. In our experience, this has been relatively harmless: most
terms in the context are there to support canonical names. Bigger terms
\emph{do} cause problems: 1. when proofs are part of mathematical objects we need to share these
proofs to allow rewriting. Moreover, it prohibits Opaque proofs and `proof irrelevance'. 2. The
projection paths may not be canonical.

@@

@@

\section{Universal algebra}\label{univ}
After having implemented concrete algebraic structures, we generalized some of our constructions
by implementing many-sorted universal algebra. A formalization of universal algebra
already exists in Coq~\cite{DBLP:conf/tphol/Capretta99}. We have conveniently redeveloped most of it
using type classes, moreover we have added a categorical framework. For instance, we show that the
term-algebra for a signature is initial. We defined equational theories and their varieties:
the categories of their models. A quote function helps to translate a statement for an instance of
a type class to a statement in the equational logic. We are currently pursuing quoting with
unification~\cite{Hints} using type classes.

We plan to include the first homomorphism theorem and perhaps the soundness theorem for equational
logic

%There is a slight problem when defining the forgetful functor from the category of
%$\Sigma$-algebras to Sets. This problem is caused by our use of the category of all
%categories\footnote{Perhaps we should have used higher categories for this.}.
%We conjecture that this problem disappears when Coq's universe polymorphism is extended to
%definitions.

\section{Interfaces using category theory}\label{interfaces}\label{modul}
The current work was stimulated by our desire~\cite{Riemann} to move between several
implementations in a flexible way. We use ad-hoc polymorphism to capture the natural
numbers without referring to their actual implementation, we have
characterized them as the initial object in the category of semirings and derived many of its
properties from this interface. Unary, binary and machine numbers are
instances of this interface, so we can directly apply these results to them.

We encourage more efficient implementations by assigning the default implementation a
low priority. For example, the distance function on the natural numbers, which is derived from its
semiring structure, is assigned priority 10.
\begin{code}
  Global Program Instance: NatDistance N | 10 := ...
\end{code}
\vspace*{-5ex}
\section{Conclusions}
Telescopes have been criticized~\cite{Packed} for the lack of multiple inheritance and
the efficiency penalty of a long chain of coercion projections. Packed classes~\cite{Packed} provide
a solution to these problems. We provide an alternative solution. We have already
discussed multiple inheritance in Section~\ref{hierarchy}. Moreover, by exposing the
carrier we avoid the chains of projections.

Packed classes use canonical structures for the algebraic hierarchy. Both canonical structures and
type classes may be seen as instances of hints in unification~\cite{Hints}. Some uses of canonical
structures can be replaced by type class instances. The user manual (2.7.15) uses canonical
structures to derive the setoid equality on the natural
numbers in the following example |(S x)==(S y)|. We have conveniently reimplemented this with
type classes with superior proof terms. Canonical structures give 
\begin{code}
@equiv(Build_Setoid nat (@eq nat)(@eq_equivalence nat))(S x)(S y)
\end{code}
